---
title: "GDP visualizations"
date: "12/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
library(dplyr)
library(ggplot2)
library(magrittr)
```

```{r}
gapminder <- read.csv("~/Data/GP_population_5Y.csv")
View(gapminder)
```

#Looking forward to EDA III: ggplot2
```{r starts.with}
# Get the start letter of each country
starts.with <- substr(gapminder$country, start = 1, stop = 1)
unique(starts.with) #observing the unique letters

# Filter countries that start with "A" or "Z"
az.countries <- gapminder[starts.with %in% c("A", "Z"), ]

# Make the plot
ggplot(data = az.countries, aes(x = year, y = lifeExp)) +
  geom_line() + 
  facet_wrap( ~country)

ggplot(data = az.countries, aes(x = year, y = lifeExp, color = continent)) +
  geom_line() + 
  facet_wrap( ~country)
```

#using the pipe operator - chaining method
```{r gpplot2}
gapminder %>%
   # Get the start letter of each country
   mutate(startsWith = substr(country, start = 1, stop = 1)) %>%
   # Filter countries that start with "A" or "Z"
   filter(startsWith %in% c("A", "Z")) %>%
   # Make the plot
   ggplot(aes(x = year, y = lifeExp, color = continent)) + #plot
   geom_line() + #geometry
   facet_wrap( ~country) #facet
```

```{r}
gapminder %>%
    # Filter countries that start with "A" or "Z"
	filter(substr(country, start = 1, stop = 1) %in% c("A", "Z")) %>%
	# Make the plot
	ggplot(aes(x = year, y = lifeExp, color = continent)) + #plot
	geom_line() + #geometry
	facet_wrap( ~country) #facet

```

```{r}
library(ggplot2)
library(plotly)
library(gapminder)

p <- gapminder %>%
  filter(year==1977, continent=='Africa', lifeExp >50) %>%
  ggplot(aes(gdpPercap, lifeExp, size = pop, group=country)) +
  geom_point() +
  theme_bw()

ggplotly(p)
```

#deeper dive into specific countries

#Selecting Ghana and Senegal 
```{r filter}
g3<-gapminder %>% 
  filter(country=='Ghana'| country=='Senegal') 
 
g4<-gapminder %>% 
  filter(country %in% c('Ghana','Senegal'))
```

```{r}
ggplot(data=g3, aes(x=year, y=gdpPercap, color=country))+
  geom_line(size=2.5)

ggplot(g4, # Change colors of lines by group
       aes(x = year, y = gdpPercap, group = country)) +
  geom_point(aes(col = country),size=1.5)
```

```{r}
ggplot(data=g3, aes(x=year, y=gdpPercap,color=country)) +
  geom_line(size=1) + scale_y_continuous(breaks=c(0,250,500,750,1000,1250,1500,1750))
```

```{r}
ggplot(data=g3, aes(x=year, y=gdpPercap, group=country, color=country)) +
  geom_line(size=1.5) +                                    xlab("Year")+ 
  ylab("GDP/Capita") +
  ggtitle("Comparing Per Capita GDP of Ghana and Senegal over the Years")
```

```{r}
ggplot(data=g3, aes(x=year, y=gdpPercap, color=country)) +
  geom_line(size=1.5)+ 
  xlab("Year")+ 
  ylab("GDP/Capita")+
  ggtitle("Comparing Per Capita GDP of Ghana and Senegal over the Years")+
   theme(plot.title= element_text(face = "bold", size = 9, hjust = 0.5), #bring the title to the middle
        axis.text.x   = element_text(size=12),
        axis.text.y   = element_text(size=13),
        axis.title.x  = element_text(size=14),
        axis.title.y  = element_text(size=15),
        legend.text   = element_text(size=16),
        legend.title  = element_text(size=17),
        panel.background = element_rect(fill =         "grey5"),
        panel.grid.major.y = element_line(colour = "grey10"))
```

```{r}
ggplot(data=g3, aes(x=year, y=gdpPercap, group=country,color=country)) +
  geom_line(size=1.5) +                           xlab("Year") + 
  ylab("GDP/Capita")+
  ggtitle("Comparing Per Capita GDP of Ghana and Senegal over the Years") +
  theme(plot.title= element_text(face = "bold",size = 10, hjust = 0.5), #bring the title to the middle
        axis.text.x   = element_text(size=14),
        axis.text.y   = element_text(size=14),
        axis.title.x  = element_text(size=14),
        axis.title.y  = element_text(size=14),
        legend.text   = element_text(size=14),
        legend.title  = element_text(size=14),
        panel.background = element_rect(fill =         "white"),
        panel.grid.major.y = element_line(colour = "grey10"))+ 
  scale_color_manual(values=c("#CC6666", "#9999CC"))
```

#Now, try to understand the story hidden in the data. 
```{r}
ggplot(data=g3, aes(x=year, y=gdpPercap, group=country,color=country)) +
  geom_line(size=1.5)+                          
  xlab("Year")+ 
  ylab("GDP/Capita")+
  ggtitle("Comparing Per Capita GDP of Ghana and Senegal over the Years")+
  theme(plot.title= element_text(face = "bold",size = 11, hjust = 0.5), #bring the title to the middle
        axis.text.x   = element_text(size=14),
        axis.text.y   = element_text(size=14),
        axis.title.x  = element_text(size=14),
        axis.title.y  = element_text(size=14),
        legend.text   = element_text(size=14),
        legend.title  = element_text(size=14),
        panel.background = element_rect(fill =         "white"),
        panel.grid.major.y = element_line(colour = "grey10"))+ 
  scale_color_manual(values=c("#CC6666", "#9999CC")) +
  geom_vline(xintercept = 1987, linetype="dashed", color = "red", size=1.0) +
  annotate(geom="text", x=1988, y=2000, label="downturn", color="red",size=6)

```

#Ghana was a country of immigration in the early years after its 1957 independence
#During the 1970s, severe drought and an economic downturn transformed Ghana into a country of emigration


#Senegal and French Sudan were merged and granted independence in 1960 as the Mali Federation. The union broke up after only a few months
#Senegal is one of the most stable democracies in Africa and has a long history of participating in international peacekeeping and regional mediation. 
#Senegal’s high total fertility rate of almost 4.5 children per woman continues to bolster the country’s large youth cohort – more than 60% of the population is under the age of 25.